Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2023-01-10 edubots_surveys.Rnw
Review RQ1 - first pars

% Describe the survey and population
We collected responses from 282 teachers from Spain and Spanish-speaking
countries for \textit{survey1} and 205 for \textit{survey2}. The two forms were
open for approximately the same amount of time, around two months. The sample
includes mostly teachers from university, although about 32\% and 8\% are from
non-universitary tertiary education in the first and second survey respectively. Regarding
gender, 61\% of the teachers were male and 32\% female, while approximately 5\%
chose not to disclose it. Finally, the responses are more or less equally
distributed according to the teaching experience, showing 24\% of responses from
teachers with 5 or less years of experience, 26\% for 6-15 years of experience,
30\% for 16-25, and 20\% for teachers with more than 25 years of experience.

We will try to find the answers to the four research questions next, by
analyzing the response to the different survey questions.

@@

Group messaging apps have become a powerful tool for \textcolor{black}{communicating} with
students. Also, many of them have built-in chatbots to enhance the learning
process. After an initial analysis to the answers to the first survey, we
realized that there were some prior issues mainly related to the adoption of a
technology, \textcolor{black}{i.e. chatbots}, that generally piggybacks on another, i.e. messaging
applications. Generally, in a technology adoption model, perceived ease of use and
perceived usefulness are essential. But it is going to be very difficult for
teachers to find {\em chatbots} easy to use if they, previously, do not
consider, or simply use, the messaging tools to which they are attached.

This is why for answering RQ1, teachers were queried about whether they use
messaging apps in their classes, specifically, Telegram, WhatsApp, Slack (an
application used mainly in IT departments and software development), or any
other messaging app, and whether they use messaging apps provided by their
academic institution (see Table \ref{tab:use_messapps}).
Overall, the majority of the teachers responded that they do use messaging apps
in their classrooms, from which apps (or, more probably, messaging solutions) provided by the academic institution
(n=159, 56\%) and WhatsApp (n=124, 44.0\%) were the most common responses. Only
19 teachers (6.7\%) replied that they do not use any messaging app in their
class.
%
\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5}
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps to assist the learning process.}
\label{tab:use_messapps}
\end{table}

% Fran: I think figures with data about the use of messaging apps should come here. Then, we say we also asked about changes after covid and include data about that
Results in Figs. \ref{figure:provided*chatbots_use},
\ref{figure:whatsapp*chatbots_use}, \ref{figure:telegram*chatbots_use} do not
show significant differences in the use of instant messaging apps between
teachers from University and vocational education. In general, most teachers
prefer messaging apps provided by their own institutions (we also show the
results for the two most popular messaging app platforms: WhatsApp and
Telegram). With respect to specific disciplines, Engineering and Technology
teachers are more active in their use, but the number of teachers from Humanities
who answered they used these apps in their classes is also remarkable
(around 60\% use the apps provided by their institutions).

@@

However, younger teachers also use Telegram with more than 25\% responding they do, a percentage

@@

Some of the questions in the survey were focused on the impact of the COVID19
pandemic between the 2020 and 2021 academic years in the teachers' attitudes towards the use of
instant messaging apps in their class. Our main intention in this case was to
assess whether a crisis will bring about some kind of change in the use of
tools. The 282 answers are summarized in

@@

pandemic lockdowns that forced students and educators to use remote education
schemes. Moreover, approximately 15\% of them switched their messaging app for

@@

Furthermore, a chi-square test of independence was performed in order to examine the relation between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the COVID-19 pandemic. The relation between the latter variable and the instructors’ sector was significant, $X^{2}_{(4, N = 282)} = 9.598$, $p =0.048$. Table \ref{tab:sector*covid} shows the frequencies that are cross-tabulated vs. changes undertaken after the COVID pandemic. This finding suggests that how teachers used messaging apps during the pandemic, specifically whether they changed their habits for teaching purposes, was related to the sector (university vs vocational) in which they were working.
%
\begin{table}[h!tbp]
\begin{tabular}{llccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Sector} & Total \\ \cline{3-5}
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}University \\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Vocational\\ (f)\end{tabular} &  \\ \cline{2-5}
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer messaging \\ app now\end{tabular} & 27 & 13 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn’t use any \\ messaging app before, but \\ do use it now\end{tabular} & 31 & 13 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before and \\ use it now\end{tabular} & 105 & 60 & 165 \\
 & \begin{tabular}[c]{@{}l@{}}No, I didn’t use it before and \\ don’t use it now\end{tabular} & 22 & 2 & 24 \\
Total &  & 193 & 89 & 282 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Sector * post-covid changes cross-tabulation}
\label{tab:sector*covid}
\end{table}

>>>>

% Describe the survey and population
We collected responses from 282 teachers from Spain and Spanish-speaking
countries for \textit{survey1} and 205 for \textit{survey2}. The two forms were
open for approximately the same amount of time, around two months. The sample
includes mostly university teachers, although about 32\% and 8\% are from
non-universitary tertiary education in the first and second survey respectively. Regarding
gender, 61\% of teachers were male and 32\% female, while approximately 5\%
chose not to disclose it. Finally, the responses are more or less equally
distributed according to the teaching experience, showing 24\% of responses from
teachers with 5 or less years of experience, 26\% for 6-15 years of experience,
30\% for 16-25, and 20\% for teachers with more than 25 years of experience.

We will try to find the answers to the four research questions next, by
analyzing the responses to the different survey questions.

@@

Group messaging apps have become a powerful tool for communicating with
students. Also, many of them have embedded chatbots to enhance the learning
process. After an initial analysis of the answers to the first survey, we
realized that there were some prior issues mainly related to the adoption of a
technology, i.e. chatbots, that generally piggybacks on another, i.e. messaging
applications. Generally, in a technology adoption model, perceived ease of use and
perceived usefulness are essential. But it will be very difficult for
teachers to find {\em chatbots} easy to use if they do not already
consider or simply use, the messaging tools to which they are attached.

This is why, in order to answer RQ1, teachers were queried about whether they use
messaging apps in their classes, specifically Telegram, WhatsApp, Slack (an
application used mainly in IT departments and software development), or any
other messaging app. They were also asked if they use messaging apps provided by their
academic institution (see Table \ref{tab:use_messapps}).
Overall, the majority of teachers responded that they do use messaging apps
in their classrooms. The most common respones were messaging apps (or, more probably, messaging solutions) provided by their academic institutions (n=159, 56\%) and WhatsApp (n=124, 44.0\%). Only
19 teachers (6.7\%) replied that they do not use any messaging app in their
class.
%
\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5}
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps to assist the learning process.}
\label{tab:use_messapps}
\end{table}

% Fran: I think figures with data about the use of messaging apps should come here. Then, we say we also asked about changes after covid and include data about that
Results in Figs. \ref{figure:provided*chatbots_use},
\ref{figure:whatsapp*chatbots_use}, \ref{figure:telegram*chatbots_use} do not
show significant differences in the use of instant messaging apps between
teachers from universities and vocational education institutions. In general, most teachers
prefer messaging apps provided by their own institutions (we also show the
results for the two most popular messaging app platforms: WhatsApp and
Telegram). With respect to specific disciplines, Engineering and Technology
teachers are more active in their use, but it is also remarkable the number of teachers from Humanities
who answered they used these apps in their classes (around 60\% use the apps provided by their institutions).

@@

However, younger teachers also use Telegram with more than 25\% responding that they do, a percentage

@@

Some of the questions in the survey were focused on the impact of the COVID-19
pandemic between the 2020 and 2021 academic years on the teachers' attitudes towards the use of
instant messaging apps in their classes. Our main intention in this case was to
assess whether a crisis will bring about a change in the use of
these tools. The 282 answers are summarized in

@@

pandemic lockdowns that forced students and educators to use remote education. 
Additionally, approximately 15\% of them switched their messaging app for

@@

Furthermore, a chi-square test of independence was performed in order to examine the relationship between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the COVID-19 pandemic. The relationship between the latter variable and the instructors’ sector was significant, $X^{2}_{(4, N = 282)} = 9.598$, $p =0.048$. Table \ref{tab:sector*covid} shows the frequencies that are cross-tabulated versus changes undertaken after the COVID-19 pandemic. This finding suggests that how teachers used messaging apps during the pandemic, specifically whether they changed their habits for teaching purposes, was related to the sector (university vs. vocational) in which they were working.
%
\begin{table}[h!tbp]
\begin{tabular}{llccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Sector} & Total \\ \cline{3-5}
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}University \\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Vocational\\ (f)\end{tabular} &  \\ \cline{2-5}
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer messaging \\ app now\end{tabular} & 27 & 13 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn't use any \\ messaging app before, but \\ do use it now\end{tabular} & 31 & 13 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before and \\ use it now\end{tabular} & 105 & 60 & 165 \\
 & \begin{tabular}[c]{@{}l@{}}No, I didn't use it before and \\ don’t use it now\end{tabular} & 22 & 2 & 24 \\
Total &  & 193 & 89 & 282 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Sector * post-COVID-19 changes cross-tabulation}
\label{tab:sector*covid}
\end{table}
